---
title: "Plague Outbreak in India"
subtitle: "Situation Report 1"
date: "`r lubridate::today()`"
author: "Emanuel & Faikah"
output: 
 prettydoc::html_pretty:
   theme: cayman
   highlight: github
   number_sections: yes
bibliography: references.bib
link-citations: yes
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(linelist)
library(kableExtra)
library(lubridate)
library(ggthemes)
library(incidence)

population_size <- 152
```

```{r read_data, echo=FALSE, message=FALSE, warning=FALSE}

plague_clean <- read_csv("data/plague_1_update.csv") %>% 
  clean_data() %>% 
  arrange(onsetdate)

plague_hhSize <- read_csv("data/plague_1_hhSize.csv") %>% 
  select(hhID, size)

plague <- left_join(plague_clean, plague_hhSize, by = c("householdid" = "hhID")) %>% 
  mutate(agegroups = cut(age, 
                         breaks = c(0,10,20,30,40,Inf),
                         right = FALSE,
                         labels = c("<10", "[10,20)","[20,30)","[30,40)","40+")))

total_cases <-plague %>%
  tally() %>%
  as.numeric()

confirmed <- plague %>% 
  filter(status == "confirmed") %>% 
  tally() %>% 
  as.numeric()

first_case <- plague %>%
    filter(onsetdate == min(plague$onsetdate)) %>%
  select(caseid)

```

```{r case_fatality_ratio, echo=FALSE}

total_deaths <- plague %>%
  drop_na(deathdate) %>% 
  nrow()

CFR <- round(total_deaths/population_size * 100,2)

```


```{r echo=FALSE}
# time difference between onset and report date
timedifference <- plague %>% 
  mutate(timediff = difftime(reportdate, onsetdate, units = "days")) %>% 
  summarise(mean =  mean(timediff)) %>% 
  as.numeric() %>% 
  round(digits = 2)

```


```{r, echo=FALSE}

earliest_onset_date <- format(min(plague$onsetdate),"%d %B %Y")
latest_onset_date <- format(max(plague$onsetdate),"%d %B %Y")

earliest_report_date <- format(min(plague$reportdate),"%d %B %Y")
latest_report_date <- format(max(plague$reportdate),"%d %B %Y")

```
# Highlights

- **`r total_cases`** suspected cases have been reported from `r earliest_report_date` to `r latest_report_date`, with `r total_deaths` deaths
- **`r confirmed`** of **`r total_cases`** suspected cases have been confirmed.
- The case fatality rate was **`r CFR`%**
- Household 10 had the most number of confirmed cases (8 cases).
- There have been no new confirmed cases beyond `r latest_report_date`.

# Context

Northeast India is the easternmost region of India and comprises of eight states. It's a mountainous region covered by dense forests. It shares border with many countries and is affected by some form of conflict [@hassan2008breakdown].

# Surveillance

The first case (case id: `r first_case`) had an onset of illness  on `r  earliest_onset_date`. The case was reported `r  difftime(min(plague$reportdate),min(plague$onsetdate))` days later and diagnosed with pneumonic plague. Unfortunately, the patient died on the same day. 

Cases continued to be reported until `r latest_report_date`. The mean duration between reported date of onset of symptoms and reported date that of notification was `r timedifference` days. This reporting delay is likely to distort the epidemic curve and the epidemic parameters.
A total of `r total_deaths` deaths has since been reported. The majority of these cases were porbable cases. The resulting case-fatality ratio is `r CFR`%.

A total of `r total_cases` cases  were reported as suspected cases. These were reported  from `r earliest_report_date` to `r latest_report_date`. The peak occurance of cases was noted on 8 December 2019 with five cases. No new cases have been reported after `r latest_report_date`. There seems to be one late case that was reported on `r latest_report_date`. This might be due several factors: (1) the child had a long incubation period hence an outlier; (2) if the case was not an outlier, then there could be other cases between 3 January 2020 and 15 January 2020 that were not indentified. The could result in a second wave of the outbreak.

```{r, echo=FALSE, message=FALSE, fig.cap= "Figure 1: Epidemic curve of pneumonic plague cases (n=88) in northeast India between October 2019 and January 2020, by date of onset of symptoms and status"}

ggplot(data = plague, aes(x = onsetdate, fill = status))+
  geom_bar()+
  scale_x_date(date_breaks = "6 day", date_labels = "%b-%d")+
  labs(x = "Date of symptom onset",
       y = "Number of Cases") +
  theme_bw()+
  theme(plot.caption = element_text(size = 10,
                                    hjust = 0),
        axis.text.x = element_text(angle = 90))

```


```{r, echo=FALSE, warning=FALSE}
plague_summary_by_sex <-  plague %>% group_by(sex) %>% summarise(total.by.sex = n())
plague_summary <- plague %>% group_by(sex, agegroups) %>% summarise(cases = n()) %>%
  inner_join(plague_summary_by_sex, by = "sex")
```

The distribution of suspected cases among males (`r plague_summary_by_sex$total.by.sex[2]`) and females (`r plague_summary_by_sex$total.by.sex[1]`) was similar. Most notably, there were no cases for the 10-20 year age group. Most cases were reported for the 20-30 year age group with 20 and 30 cases reported by females and males, respectively. Two cases did not have information on the age of the cases.

```{r, echo=FALSE, warning=FALSE}
kable(plague_summary, col.names = c("Sex", "Age group","Number of cases", "Total"),align = "c", caption = "Table 1: Summary of cases by sex and age group")%>%
  kable_styling(bootstrap_options = "striped", full_width = T) %>%
  column_spec(1, bold = T)%>%
  collapse_rows(columns = c(1,4), valign = "middle")

```

<br>


The number of reported cases from household varied by the household sizes. Household 10 had the highest overall number of cases (eight confirmed cases).

```{r, echo=FALSE, message=FALSE, fig.cap= "Figure 2: Occurance of pneumonic plague cases per household by status"}

plague %>% 
  group_by(householdid) %>% 
  count(status) %>% 
  ggplot(aes(x = factor(householdid), y = n, fill = status))+
  geom_col()+
  theme_bw()+
  labs(x = "Household ID",
       y = "Cases")+
  theme(plot.caption = element_text(size = 10,
                                    hjust = 0))
```


# Transmissibility

```{r modelling, echo=FALSE}
# if you use the function fit_optim_split(plague_incidence), it splits the incidence automatically for you. I did daily analysis
plague_incidence <- incidence(plague$onsetdate, interval = 1) 
model_fit <- fit_optim_split(plague_incidence)
# model_fit$split
# model_fit$fit
# plot(plague_incidence, border = "white") %>% 
#   add_incidence_fit(model_fit$fit)

growth_rate <- round(model_fit$fit$before$info$r,4)
LCI_r <- round(model_fit$fit$before$info$r.conf[1],4)
UCI_r <- round(model_fit$fit$before$info$r.conf[2],4)
  
doubling_time <- round(model_fit$fit$before$info$doubling,4)
LCI_dt <- round(model_fit$fit$before$info$doubling.conf[1],4)
UCI_dt <- round(model_fit$fit$before$info$doubling.conf[2],4)
  
# stratified by sex
# plague_incidence_sex <- incidence(plague$onsetdate, interval = 1, groups = plague$sex)
# model_fit_sex <- fit_optim_split(plague_incidence_sex)
# model_fit_sex$split
# model_fit_sex$fit
# plot(plague_incidence_sex, border = "white") %>% 
#   add_incidence_fit(model_fit_sex$fit)

```

We modelled the transmissibility of plague in northeast Inda by fitting a log-linear regression model to the data. This analysis was conducted using the `incidence` package [@kamvar2019epidemic]. We found that the outbreak was growing very slowly (growth rate r was `r growth_rate` , 95% CI: [`r LCI_r`, `r UCI_r`]) and it took about two months (doubling time was `r doubling_time`, 95% CI: [`r LCI_dt`, `r UCI_dt`] days) for the incidence to double.

# References
